---
title: "SPEAR: SPEARkit Walkthrough"
author: "Jeremy Gygi"
date: "3/12/21"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPEAR_4-SPEARkit_walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

### This vignette walks through all SPEARkit functions (start with 'SPEAR.______')

##### Installing SPEAR:

Follow installation instructions in README (located [here](https://github.com/jgygi/SPEAR))

##### Required Libraries:

```{r message=FALSE, warning=FALSE}
# Required Packages for SPEAR:
library(SPEARcomplete)
library(glmnet)
library(parallel)
library(ordinalNet)

# Downstream Analysis:
library(dplyr)
library(reshape2)
library(stringr)
library(ggplot2)
library(gplots)
library(cowplot)
```

### Table of Contents:

1. **Functions that access data**
    + *Model*:
        + SPEAR.get_best_weights **NEW** (add + 1 cvsd?)
    + *Prediction*:
        + SPEAR.get_predictions **NEW** (combine SPEAR.get_cv_predictions and SPEAR.predict_new_samples)
        + SPEAR.get_cv_prediction_error (return SDs?)
        + SPEAR.get_misclassification_error **NEW** (BER error?)
    + *Downstream*:
        + SPEAR.get_factor_scores
        + SPEAR.get_factor_contributions
        + SPEAR.get_factor_features
2. **Plotting functions**


## 1) Functions that access data
### Model
#### SPEAR.get_best_weights
Returns the best weights from the models in the `SPEARobject` determined by `method`
```{r}
print("test")
```

### Prediction
### Downstream
## 2) Plotting functions







