---
title: "SPEAR: Using the new version"
author: "Jeremy Gygi"
date: "06/28/21"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPEAR_new_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

### This vignette will provide a walkthrough for using the new version of SPEAR on multi-omic data.

##### Installing SPEAR:

Follow installation instructions in README (located [here](https://github.com/jgygi/SPEAR))

##### Required Libraries:

```{r message = FALSE, warning = FALSE}
library(SPEAR)
```

### Loading the multi-omic data:

Loading your own multi-omic data:

```{r}
# Parameters:
# X - a list of matrices of omics data [subjects = rows, features = columns] (i.e. OmicsData 1 would be X[[1]], ...)
# Y - a matrix of response data [subjects = rows, response.variables = columns] (i.e. Response1 would be Y[,1], ...)
# 
# Data should be scaled/standardized (mean = 0, variance = 1) before using SPEAR:

X <- NULL # Replace with your list of matrices...
Y <- NULL # Replace with your matrix...
```

Read in simulated data...

```{r}
# Generated from the SPEAR_simulate_data.Rmd file:
path_to_sim_data <- ""
sim.data <- readRDS(paste0(path_to_sim_data, "simulated_gaussian_data.rds"))

X <- sim.data$data.tr$xlist
Y <- sim.data$data.tr$Y
```

Naming the features, subjects, and datasets (for downstream analysis):

```{r}
colnames(Y) <- "GausResponse"
rownames(Y) <- paste0("Subject", 1:nrow(Y))
names(X) <- paste0("OmicsData", 1:length(X))
for(i in 1:length(X)){
  colnames(X[[i]]) <- paste0(names(X)[i], "_feat", 1:ncol(X[[i]]))
  rownames(X[[i]]) <- paste0("Subject", 1:nrow(X[[i]]))
}
```

# Getting SPEAR to run:

## 1) Make a SPEARobject-class object:

```{r}
SPEARobj <- make.SPEARobject(X = X,
                             Y = Y,
                             family = "gaussian",
                             # options:
                             seed = 456, # set the seed (for consistent runs)
                             print_out = 10, # Print ELBO change after how many iterations? (default is 100)
                             remove.formatting = TRUE, # if TRUE, removes coloring from output (doesn't work in HTML)
                             quiet = FALSE # if TRUE, will silence all unnecessary print messages
                             )

# NOTE: you can use encoded family parameters as well (i.e. family = 0 for gaussian, 1 for binomial,
#                                                      2 for ordinal and 3 for multinomial)
```

When you initialize a SPEARobject, the `$fit` list will be `NULL`...

```{r}
names(SPEARobj$fit)
```

Some parameters are automatically initialized if you don't specify them in the `make.SPEARobject` function...

```{r}
SPEARobj$params$num_factors
```

This includes the `$params$weights` argument, which is now a matrix...

```{r}
SPEARobj$params$weights
```

We have only used one `weight` or `w` parameter up until this point. That has been renamed `w.x` (we have always had `w.y = 1`)

## 2) Running regular SPEAR (no cross validation):

```{r}
SPEARobj$run.spear()
```

This will populate the `$fit` parameter:

```{r}
names(SPEARobj$fit)
```

Right now, there aren't any of the helper functions for plotting. You can directly access all of the coefficients, however:

+ regression.coefs - post_betas

+ projection.coefs.x - post_bxs

+ projection.coefs.y - post_bys

+ nonzero.probs - post_pis

+ projection.probs - post_selections

+ marginal.probs - post_selections_marginal

+ joint.probs - post_selections_joint

+ intercepts.x - interceptsX

+ intercepts.y - interceptsY

+ type - new, either 'regular' if called from `run.spear` or 'cv' if called from `run.cv.spear`.

## 3) Running cross validation SPEAR:

NOTE: This will overwrite the results stored in the `$fit` list.

```{r}
# First, run cv spear:
SPEARobj$run.cv.spear()

# Then, run the cv evaluation function:
SPEARobj$cv.evaluate()
```

The `cv.evaluate` function does a lot of important calculations, including getting the mean cross validation error per weight:

```{r}
cvm.df <- SPEARobj$fit$cv.eval$cvm
rownames(cvm.df) <- paste0("w.idx", 1:nrow(cvm.df))
colnames(cvm.df) <- "mean.cv.error"
cvm.df
```

Now that we have a `w.x` and a `w.y` to worry about, we simplify the process by choosing a **pair of weights**, given by a weight index (`w.idx`)

We set the `w.idx` parameter either through the `set.weights(...)` function...

```{r}
SPEARobj$set.weights(w.idx = 1) # this refers to the first row of SPEARobj$params$weights...
```

... or we can set it manually.

```{r}
SPEARobj$options$current.weight.idx <- 1
```

# Save the new SPEARobject-class object:

```{r}
SPEARobj$save.model(file = "SPEARobj_test.rds")

# Can be read in with...
SPEARobj <- load.SPEARobject("SPEARobj_test.rds")
# or...
# SPEARobj <- readRDS("SPEARobj_test.rds")
```


# Analyzing the new SPEARobject-class object:

As of right now, there aren't many helper functions... below are the current ones available:

### Adding a testing dataset

If you have test data, you can add it to the SPEARobject using the `add.data(...)` function...

While providing a list of matrices for `X` is required, `Y` is optional.

*NOTE:* The column names (features) in `X` **must** match the column names of the training data...

```{r}
X.test <- sim.data$data.te$xlist
Y.test <- sim.data$data.te$Y
colnames(Y.test) <- "GausResponse"
rownames(Y.test) <- paste0("Subject", 1:nrow(Y.test))
names(X.test) <- paste0("OmicsData", 1:length(X.test))
for(i in 1:length(X.test)){
  colnames(X.test[[i]]) <- paste0(names(X.test)[i], "_feat", 1:ncol(X[[i]]))
  rownames(X.test[[i]]) <- paste0("Subject", 1:nrow(X.test[[i]]))
}

SPEARobj$add.data(X = X.test, Y = Y.test, name = "test.dataset")
```

The dataset has been stored under `$data`...

```{r}
names(SPEARobj$data)
```

Many of the functions below support choosing which dataset to use.

### Factor Scores:

Return a matrix of N X K (num. samples x num. factors). Can choose to get in.sample (`cv = FALSE`) or cross-validated scores (`cv = TRUE`). For most purposes, leave `cv = FALSE`.

```{r fig.width = 4, fig.height = 5, fig.align = "center"}
factor.scores = SPEARobj$get.factor.scores()
factor.scores.cv = SPEARobj$get.factor.scores(cv = TRUE) # cv defaults to FALSE above
heatmap(factor.scores)
```

You can get new factor scores for a dataset added via the `add.data` function...

```{r fig.width = 4, fig.height = 5, fig.align = "center"}
factor.scores = SPEARobj$get.factor.scores(data = "test.dataset")
factor.scores.cv = SPEARobj$get.factor.scores(data = "test.dataset", cv = TRUE) # cv defaults to FALSE above
heatmap(factor.scores)
```

### Predictions:

This will return the predictions (in a vector if `family = "gaussian"`, and in a named list otherwise)...

```{r fig.width = 6, fig.height = 5}
preds <- SPEARobj$get.predictions() # Defaults to use CV factors (cv = TRUE) and data = "train"
true <- SPEARobj$data$train$Y
plot(true ~ preds)
```

You can also use datasets from the `add.data` function here...

```{r fig.width = 6, fig.height = 5}
preds <- SPEARobj$get.predictions(data = "test.dataset") # Defaults to use CV factors (cv = TRUE)
true <- SPEARobj$data$test.dataset$Y
plot(true ~ preds)
```

### Factor contributions:

You can get the factor contributions for X and Y:

```{r}
factor.contributions <- SPEARobj$get.contributions() # defaults to return X and Y, use do.X = FALSE for only Y...
factor.contributions
```

Notice that Factor 1 and Factor 2 contribute the most to the prediction of Y

### Features:

Get a data.table of features using a cutoff:

```{r}
features <- SPEARobj$get.features() # defaults to coefficient.cutoff = 0 and probability.cutoff = .95
nrow(features)

# To get ALL features, just set both cutoffs to 0:
all.features <- SPEARobj$get.features(coefficient.cutoff = 0,
                                      probability.cutoff = 0)
nrow(all.features)

# To add columns including correlation with factor scores, use the correlation argument:
features.with.correlation <- SPEARobj$get.features(correlation = "pearson")

features.with.correlation[1:10,]

# Get features for a specific Factor / Dataset:
features.factor1.dataset3 = SPEARobj$get.features(factors = 1, datasets = "OmicsData3")
head(features.factor1.dataset3)
```


